---
title: "A Constrained Optimization Formulation of the Indefinite Procrustes Problem"
author: "Joshua Agterberg"
date: ""
output: html_document
header-includes: 
  - \usepackage{amsmath}
bibliography: procrustes.bib
---
```{r,include=FALSE}
library(Matrix)
library(irlba)
library(ggplot2)
library(devtools)
library(knitr)
library(rgl)
knit_hooks$set(webgl = hook_webgl)
knitr::opts_chunk$set( fig.align="center")
```
Document explaining the code for indefinite procrustes analysis.  First, the indefinite procrustes solution code comes from the following place:
```{r,message=FALSE}
devtools::source_url(url="https://raw.githubusercontent.com/jagterberg/indefinite_procrustes/master/indefinite_procrustes.R")
```
```{r,include=FALSE}
simulateAdjacency <- function(B = B_mat,n = 99,b= c(rep(1,n/3),rep(2,n/3),rep(3,n/3))) {
  A <- matrix(0,n,n)
  for (i in c(1:n)) {
    j <- n
    while (j > i) {
      A[i,j] <- rbinom(n=1,size=1,p=B[b[i],b[j]])
      A[j,i] <- A[i,j]
      j <- j -1
    }
  }
  
  return(A)
}
procrustes_plot <- function(True,X,Y,which.inds = c(1,2),names = NULL,truth.ind = NULL
                            , title = c('Spectral Embedding of Indefinite Matrix after \nApplying Indefinite Procrustes Analysis')
                            ,transparency = .2,true.size = 4) {
  n <- dim(X)[1]
  if (is.null(names)) {
    names <- factor(c(rep('Rotated',n),rep('Original',n),rep('True',n)),
                    levels=c('Rotated','Original','True'))
  }
  
  if (is.null(truth.ind)) {
    truth.ind <- factor(c(rep(0,2*n),rep(1,n)),levels = c("0","1"))
  }
  
  dat <- data.frame(X1 = c(X[,which.inds[1]],Y[,which.inds[1]],True[,which.inds[1]])
                    ,X2 =c(X[,which.inds[2]],Y[,which.inds[2]],True[,which.inds[2]])
                    ,Type = names
                    ,Indicator =truth.ind )
  names.levels <- levels(names)
  truth.ind.levels <- levels(truth.ind)
  truths.size<- c(1.5,true.size)
  truths.alpha <- c(transparency,1)
  names(truths.alpha) <- truth.ind.levels
  names(truths.size) <- truth.ind.levels
  names.cols <- c('red','orange','black')
  names(names.cols) <- names.levels
  names.labels <- c('Xhat after Rotation','Xhat Original','True Values')
  names(names.labels) <- names.levels
  g <- ggplot(data=dat) + geom_point(aes(x = X1,y=X2,col = Type,size = as.factor(Indicator)
                                         ,alpha = factor(Indicator))) + 
    scale_color_manual(values = names.cols,
                       labels= names.labels
    ) + guides(colour = guide_legend('')) +
    scale_alpha_manual(values = truths.alpha,guide=FALSE) +
    scale_size_manual(values = truths.size
                      ,guide = FALSE) +
    ggtitle(title)
  
  print(g)
}


#' Function to create the true latent position vectors from an SBM 
#' with block probabilities B and latent assignment vector b
get_true_X <- function(n,B,b= c(rep(1,n/3),rep(2,n/3),rep(3,n/3))) {
  X <- matrix(0,n,n)
  for (i in c(1:n)) {
    j <- n
    while (j >= i) {
      X[i,j] <- B[b[i],b[j]]
      X[j,i] <- X[i,j]
      j <- j - 1
    }
  }
  d <- length(unique(b))
  X_true <- irlba(X,d)
  X_true <- X_true$u[,c(1:d)] %*% diag(X_true$d[c(1:d)])^(.5)
  return(X_true)
}


#' Function to plot procrustes for two block model:
procrustes.simulation2 <- function(seed,n,p = .2,q=.7,r=.2) {
  set.seed(seed)
  B_mat <- matrix(c(p,q,q,r),2,2)
  b <- c(rep(1,n/2),rep(2,n/2))
  A <- simulateAdjacency(n=n,B= B_mat,b=b)
  Xtrue <- get_true_X(n,B = B_mat,b=b)
  A_svd <- irlba(A,2)
  Xhat <- A_svd$u[,c(1:2)] %*% diag(A_svd$d[c(1:2)])^(.5)
  solution <- solve_procrustes_problem(X=Xhat,Y=Xtrue,p=1,q=1)
  
  Xhat_rotated <- Xhat %*% convertToMatrix(solution$par)
  
  procrustes_plot(True = Xtrue,Y = Xhat,X = Xhat_rotated,transparency = .3,true.size= 3)
}
```
Here are the results from simulating an adjacency matrix from the block probability matrix $B = \begin{pmatrix} .2 & .7 \\ .7 & .2 \end{pmatrix}$ with $\frac{n}{2}$ vertices in each block (here, $n = 1000$).  The printed results below on are the output from each iteration of the `alabama::auglag()` R function, which runs the Augmented Lagrangian on our indefinite procrustes problem.  The code is attempting to solve
$$\underset{{U \in \mathcal{O}_{p,q}}}{\operatorname{argmin}} ||XU - Y||_F^2 $$
numerically (where $X,Y \in \mathbb{R}^{n \times d}$, and $p+q = d$),  although this is reformulated according to that given in @indefinite_procrustes.  The plot examines the corresponding plotted point clouds in $\mathbb{R}^2$ both before and after transformation.  

Note: the solution can be read off directly from the output (the par argument) by simply thinking of it as a $2 \times 2$ matrix (filling in each entry by row first).
```{r}
procrustes.simulation2(seed=24,n=1000)
```
For many of the simulations on the indefinite 2-block model, the optimal matrix is very close to either the identity or the negative identity (or some combination thereof).  Since we are interested in performance more generally,  we will also examine a 3-block model with $B := \begin{pmatrix} .6 & .9 &.9 \\ .9 & .6 &.9 \\ .9 &.9 &.3 \end{pmatrix}$, with $\frac{n}{3}$ vertices in each block. Included are: The results from the `auglag()` function, an interactive 3D plot to examine the point clouds in $\mathbb{R}^3$, and individual plots along each axis.
```{r,include=FALSE}
procrustes.simulation3 <- function(seed,n
                                   ,B_mat =  matrix(c(.6,.9,.9,.9,.6,.9,.9,.9,.3),3,3)
                                   ,dims = c(1,2)) {
  set.seed(seed)
  b <- c(rep(1,n/3),rep(2,n/3),rep(3,n/3))
  A <- simulateAdjacency(n=n,B= B_mat,b=b)
  Xtrue <- get_true_X(n,B = B_mat,b=b)
  A_svd <- irlba(A,3)
  Xhat <- A_svd$u[,c(1:3)] %*% diag(A_svd$d[c(1:3)])^(.5)
  solution <- solve_procrustes_problem(X=Xhat,Y=Xtrue,p=1,q=2)
  
  Xhat_rotated <- Xhat %*% convertToMatrix(solution$par)
  
  title <- paste0("Spectral Embedding and Rotated Spectral Embedding of Indefinite Matrix\n(on x = dimension ",dims[1],", y = dimension ",dims[2],")")
  procrustes_plot(True = Xtrue[,dims],Y = Xhat[,dims],X = Xhat_rotated[,dims]
                  ,transparency = .3,true.size= 3
                  ,title =title)
  
}
seed = 100
procrustes.simulation3(seed=seed,n=666,dims = c(1,2))
procrustes.simulation3(seed=seed,n=666,dims = c(1,3))
procrustes.simulation3(seed=seed,n=666,dims = c(2,3))


seed =300

procrustes.simulation3(seed=seed,n=666,dims = c(1,2))
procrustes.simulation3(seed=seed,n=666,dims = c(1,3))
procrustes.simulation3(seed=seed,n=666,dims = c(2,3))



procrustes_plot3D <- function(True,X,Y,names = NULL,truth.ind = NULL
                            , title = c('Spectral Embedding and Rotated Spectral Embedding of Indefinite Matrix')
                            ,transparency = .2,true.size = 4) {
  n <- dim(X)[1]
  if (is.null(names)) {
    names <- factor(c(rep('Rotated',n),rep('Original',n),rep('True',n)),
                    levels=c('Rotated','Original','True'))
  }
  
  if (is.null(truth.ind)) {
    truth.ind <- factor(c(rep(0,2*n),rep(1,n)),levels = c("0","1"))
  }
  
  dat <- data.frame(X1 = c(X[,1],Y[,1],True[,1])
                    ,X2 =c(X[,2],Y[,2],True[,2])
                    ,X3 = c(X[,3],Y[,3],True[,3])
                    ,Type = names
                    ,Indicator =truth.ind )
 
  dat$Type2 <- ifelse(dat$Type == "Rotated","red",
                     ifelse(dat$Type == "True","black","lightblue"))
  title <- "Spectral Embedding and Rotated Spectral Embedding of Indefinite Matrix"
  title2 <- "Light Blue: Non-Rotated, Red: Rotated, Black: True"
  with(dat,plot3d(X1,X2,X3,col = dat$Type2,size = 0,main=title,sub=title2,box=FALSE,axes=TRUE))
 
  irisList <- split(dat,dat$Indicator)
  
  # Setup the plot

  # Use a separate call to points3d() to plot points of each size
  for(i in seq_along(irisList)) {
    if (irisList[[i]]$Indicator[1] == "1") {
      j <- 10
    
    } else {
      j <- 2
    }
    with(irisList[[i]], points3d(X1,X2,X3, col=Type2, size=j))
  }
  

  
}

procrustes.simulation3D <- function(seed,n
                                    ,B_mat =  matrix(c(.6,.9,.9,.9,.6,.9,.9,.9,.3),3,3)
                                    ) {
  set.seed(seed)
  b <- c(rep(1,n/3),rep(2,n/3),rep(3,n/3))
  A <- simulateAdjacency(n=n,B= B_mat,b=b)
  Xtrue <- get_true_X(n,B = B_mat,b=b)
  A_svd <- irlba(A,3)
  Xhat <- A_svd$u[,c(1:3)] %*% diag(A_svd$d[c(1:3)])^(.5)
  solution <- solve_procrustes_problem(X=Xhat,Y=Xtrue,p=1,q=2)

  Xhat_rotated <- Xhat %*% convertToMatrix(solution$par)
  #print(paste0("Final solution: U = ", solution$par))
  #title <- paste0("Procrustes Analysis For D=3 \n(on x = dimension ",dims[1],", y = dimension ",dims[2],")")
  
  procrustes_plot3D(True = Xtrue,Y = Xhat,X = Xhat_rotated
                    ,transparency = .3,true.size= 3
                    ,title =title)
}
```
```{r,echo=FALSE,webgl=TRUE}
procrustes.simulation3D(seed=100,n=666)
```
```{r, echo=FALSE,results='hide',fig.keep='all'}
seed =100
procrustes.simulation3(seed=seed,n=666,dims = c(1,2))
procrustes.simulation3(seed=seed,n=666,dims = c(1,3))
procrustes.simulation3(seed=seed,n=666,dims = c(2,3))
```
